Outline flowCore paper

### 2008-04-08 paper/outline.tex
Added a few sentences about Gating-ML and fixed some typos in the text.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@30924 bc3139a8-67e5-0310-9ffc-ced21a209358

Traditionally, flow cytometry (FC) has been a tube-based technique
limited to small-scale laboratory and clinical studies.  High
throughput methods for flow cytometry (FC-HCS) have recently been
developed and are now used in both basic and clinical research. Today,
the need for flexible and well structured computational tools to
efficiently handle and analyze FC-HCS has increased
considerably. Indeed, the amount of information generated by these
technologies must be stored, managed, and needs to be summarized in
order to be accessible to the researcher. However, the absence of any
shared research platform which bioinformaticians, computer scientists,
and statisticians can utilze to develop novel or standard methods for
flow cytometric has become problematic. We believe that the open
source statistical software R in conjuction with the Bioconductor
Project can fill this gap.  In this paper, we propose R data
structures to handle flow cytometry data through the main steps of
importing, storing, assessing and preprocessing data from FC-HCS
experiments. These structures, along with a range of specialized
methods for compensation, transformation, gating and additional
preprocessing steps, are implemented in the newly developed
Bioconductor package flowCore.  We envision flowCore to be a solid
foundation for researchers to efficiently handle FC-HCS data and for
future development of new methods and tools for their coherent
analysis.

\subsection{R and Bioconductor}
R is a solid statistical programing environment, particularly geared
towards the analysis of large numeric datasets typical to arise from
high throughput experiments such as microarray gene expression anlysis
or FC-HCS. Today, the R open source project offers a wide range of
statistical and visualization methods developed for various fields of
application. In addition, R is a research platform which
bioinformaticians, computer scientists, and statisticians routinely
use to develop novel tools and methodology for statistical data
analyis. For biological applications, the Bioconductor project has
become the standard toolset \citep{gentleman2006bos} to process data
spanning a diversity of research fields from genomics to proteomics to
complex cell biology. This shared platform proved to be especially
beneficial for the development of analyis routines for microaray gene
expression data, and by now many of these innovations have been
incoorporated into comercial software products, thus greatly improving
the quality of gene expression data analysis. Shared data structures
have been shown to be essential for this kind of collaborative effort.

\subsection{Existing data standards and conventions}
Currently, data from flow cytometry experiments are stored in single
files following a well-defined data standard (FCS). This standard
tries to capture as much of the accompanying meta information as
possible, most of which is produced by the measurement instrument or
has to be supplied manually by the operator upon experiment
setup. However, the recent developments in high-throughput flow
cytometry are shifting the focus of interest away from single-tube
based measurements towards large and complex experimental designs with
dozens of co-variates and influencing factors. Self-contained data
structures are a prerequesite to allow for coordinated analysis of
such data sets. Most of the currently available software solutions
offer only limited support for such structures, or make use of
containers like XML or binary storage formats that are designed
specifically for the mostly GUI-driven user interfaces and hence are
not easily amenable to programatical access. In addition, the
closed-source nature of most of these products makes it virtually
impossible to connect and integrated these data.

Traditionally, the majority of flow cytometry experiments have been
analysed manually by direct data inspection or very basic comparisons
of summary statistics. We believe that these approaches do not give
justice to the highly complex nature of FC data, essentially
discarding many of the fundamental aspects such as data distribution
or high-dimensionality. In addition, the subjective character of
manual analyses are a major obstacle for reproducibility. For FC-HCS
data, unassisted manual inspection is no longer feasible, and robust
statistical methods need to be developed to point the investigator to
the intersting aspects of the data or to potential problems. While the
expert knowledge of immunologists and researchers remains crucial for
the understanding of FC data, we believe that tight colaboration
with other reserach fields such as statistics and computer science can
greatly improve the relevance of flow cytomtry in today's
high-throughput biological sciences.

\textit{How does the flow data world look like? Other flow software
  (what other software are handling multiple FCS files at
  all). Somebody more familiar with flowJo and friends should add to
  this... Also somebody should fill in a couple of sentences about the
  gating-ML etc. stuff here}

The flowCore framework presented here offers import functionality for
both the raw data FCS files along with the complete set of
file-specific meta information as well as all the information that is
stored in one of the gating-, compensation- or transformation-ML
standards. Thus, it is possible to intergrate flowCore in existing
work flows and to communicate with virtually every other FC tool that
adheres to these standards.

@@

\section{Basic data structures}
\subsection{flowFrame}
flowCore's primary task is the representation and basic manipulation
of flow cytometry (or similar) data. This is accomplished through a
data model very close to that adopted by other Bioconductor packages
used for gene expression analysis and thus familiar to most
Bioconductor users. Similar to any other flow tool, the basic unit of
organisation in flowCore is a collection of events. We call the
structure that hold this data a flowFrame. All relevant information
for the collection are stored along with the matrix of raw data values
and can be accessed programatically. Most prominently, these are
descriptors of the stains used in the experiment and the respective
measurement channels, information about compensation performed at the
instrument side and any additional keywords the user deems to be
important to annotate the data. A number of quality checks are
performed during the creation of a flowFrame to ensure data
integrety. There is an extended interface that allows for the
standardized interaction with the flow data, thus enabling the
development of more specialized tools for specific tasks of the data
analysis process.

\subsection{flowSet}
In high-throughput flow cytometry, many of the analysis tasks need to
be performed consistently across multiple flow measurements, hence we
introduce the concept of a collection of flowFrames called a
flowSet. flowSets store the relevant information associated with each
individual frame such as descriptions of the samples or the treatment
to which a sample was subjected, and they manage the consistent
application of operations on the individual flowFrames. The burden of
keeping score of the annotation is shifted from the user to the
infrastructure, thus reducing the risk of mixups and errors that are
all too common when dealing with complex data in the ususal way of
more or less ad hoc spreadsheats that lack a controlled structure. In
general, there is no native ordering to a flowSet, which in a way
reflects the still predominant application of tube-based experiments
rather than experiments following a plate-design. However, the flowSet
structure can be readily extended to also accomodate such spacial
information \textit{(reference)}

@@

\section{Standard flow operations}
The basic operations in flow cytometry analysis are almost always the
same: the data need to be compensated (if not already done on the
instrument side), often transformed and sup-population of interest
need to be selected based on a set of (predominantly sequential)
gates. All software solutions for flow data analysis offer more or
less elaborate support for these kinds of operation, most of them in
an interactive, GUI-driven spirit. The approach we have taken in
flowCore is to abstractly describe these operations and build a set of
tools to perform them on both flowFrames as well as on entire
flowSets. While transformation and to a certain extent also
compensation are fairly routine operations with only limited potential
for improvement, being able to implement new methodology for gating of
flow cytometry data and extending the capabilies of flowCore through
the usual ways of object oriented programming is a feature, that
clearly sets our framework apart from all the other flow toolsets
around. By factoring out as much of the bookkeeping as possible,
authors can focus on the actual operations rather then having to deal
with the teadious details of data integration and access. Third party
methods implemented by others can act on their own as first-class
citizens in the analysis, without breaking the workflow or the basic
infrastructure. This design allows for the straight forward extension
of flowCore's capabilities and already fostered the development of a
number of valuable improvements. \textit{(cite stuff here)}

@@

\subsection{gating}
In flowCore, gates are represented as individual objects that can be
extended in an object oriented manner. All basic gate types such as
rectangular gates, ellipses and polygons are implemented as part of
the base framework. In addition, we introduce the notion of
data-driven gates, for which the necessary parameters are computed
based on the properties of the underlying data, for instance
distribution or density estimates. The ability to programatically
access gates is a prerequesite for automated or semi-automated
gating. By utilizing a unified interface for all different types of
gates, the user is able to subset data sets as well as create summary
statistics, for instance of proportions of events falling in single
gates or combinations of multiple gates. The definition of gates in
flowCore follows the proposed gating-ML standard, thus any flowCore
gating strategy can be reproduced by any other software that also
adheres to the standard.

\section{Related flow packages}
In addition to the flowCore package that offers basic infrastructure
we have implemented a range of additional Bioconductor packages that
are dedicated to more specific tasks of flow cytomtery data
analysis. The flowViz package provides sophisticated data
visualization tools that make use of multivariate trellis
plotting. These functions can be used to quickly generate customized
plots for extended cytometry data sets for both direct data inspection
and quality control. The flowQ package offers more advanced quality
assurance (QA) methodology and a framework to create interactive
reports of QA results. Most of the flowUtils content deals with data
import and output from and to flowCore and the flow-cytometry specific
standard markup languages. Finally, flowStats provides statistical
methods that are relevant in the context of flow cytometry data
analysis.

>>>>

Traditionally, flow cytometry (FCM) has been a tube-based technique
limited to small-scale laboratory and clinical studies.  High
throughput methods for flow cytometry (FC-HCS) have recently been
developed and are now used in both basic and clinical research. Today,
the need for flexible and well structured computational tools to
efficiently handle and analyze FC-HCS has increased
considerably. Indeed, the amount of information generated by these
technologies must be stored, managed, and needs to be summarized in
order to be accessible to the researcher. However, the absence of any
shared research platform, which bioinformaticians, computer scientists,
and statisticians can utilize to develop novel or standard methods for
flow cytometric analysis has become problematic. We believe that the open
source statistical software R in conjunction with the Bioconductor
Project can fill this gap.  In this paper, we propose R data
structures to handle flow cytometry data through the main steps of
importing, storing, assessing and preprocessing data from FC-HCS
experiments. These structures, along with a range of specialized
methods for compensation, transformation, gating and additional
preprocessing steps, are implemented in the newly developed
Bioconductor package flowCore.  We envision flowCore to be a solid
foundation for researchers to efficiently handle FC-HCS data and for
future development of new methods and tools for their coherent
analysis.

\subsection{R and Bioconductor}
R is a solid statistical programing environment, particularly geared
towards the analysis of large numeric datasets typical to arise from
high throughput experiments such as microarray gene expression analysis
or FC-HCS. Today, the R open source project offers a wide range of
statistical and visualization methods developed for various fields of
application. In addition, R is a research platform which
bioinformaticians, computer scientists, and statisticians routinely
use to develop novel tools and methodology for statistical data
analysis. For biological applications, the Bioconductor project has
become the standard toolset \citep{gentleman2006bos} to process data
spanning a diversity of research fields from genomics to proteomics to
complex cell biology. This shared platform proved to be especially
beneficial for the development of analysis routines for microarray gene
expression data, and by now many of these innovations have been
incorporated into commercial software products, thus greatly improving
the quality of gene expression data analysis. Shared data structures
have been shown to be essential for this kind of collaborative effort.

\subsection{Existing data standards and conventions}
Currently, data from flow cytometry experiments are stored in single
files following a well-defined data standard (FCS). This standard
tries to capture as much of the accompanying meta information as
possible, most of which is produced by the measurement instrument or
has to be supplied manually by the operator upon experiment
setup. However, the recent developments in high-throughput flow
cytometry are shifting the focus of interest away from single-tube
based measurements towards large and complex experimental designs with
dozens of co-variates and influencing factors. Self-contained data
structures are a prerequisite to allow for coordinated analysis of
such data sets. Most of the currently available software solutions
offer only limited support for such structures, or make use of
containers like XML or binary storage formats that are designed
specifically for the mostly GUI-driven user interfaces and hence are
not easily amenable to programmatical access. In addition, the
closed-source nature of most of these products makes it virtually
impossible to connect and integrated these data.

Traditionally, the majority of flow cytometry experiments have been
analysed manually by direct data inspection or very basic comparisons
of summary statistics. We believe that these approaches do not give
justice to the highly complex nature of FCM data, essentially
discarding many of the fundamental aspects such as data distribution
or high-dimensionality. In addition, the subjective character of
manual analyses are a major obstacle for reproducibility. For FC-HCS
data, unassisted manual inspection is no longer feasible, and robust
statistical methods need to be developed to point the investigator to
the interesting aspects of the data or to potential problems. While the
expert knowledge of immunologists and researchers remains crucial for
the understanding of FCM data, we believe that tight collaboration
with other research fields such as statistics and computer science can
greatly improve the relevance of flow cytometry in today's
high-throughput biological sciences.

These issues are currently being addressed by emerging new flow cytometry 
data standards developed in collaboration with the ISAC Data Standards Task Force. 
The Gating-ML Candidate Recommendation (CR) represents a proposal on how to form 
unambiguous XML-based gate definitions, which can facilitate the interchange and 
validation of data between different software packages with the potential of 
significant increase of hardware and software interoperability. Gates may be 
ordered into a hierarchical structure to describe a gating strategy. They may be 
applied on parameters as in FCS files or on transformed parameters as described 
by an explicit parameter transformation. This allows to encode a simple analytical 
work flow, which is sufficient in order to reconstruct the analysis programmatically.

\textit{How does the flow data world look like? Other flow software
  (what other software are handling multiple FCS files at
  all). Somebody more familiar with flowJo and friends should add to
  this...}

The flowCore framework presented here offers import functionality for
raw data FCS files along with a complete set of file-specific meta information.
Moreover, it is one of the very first software implementations of the Gating-ML CR, 
which opens the possibility to integrate flowCore in existing work flows and to 
communicate with virtually every other FCM tool that adheres to this standard.

@@

\section{Basic data structures}
\subsection{flowFrame}
flowCore's primary task is the representation and basic manipulation
of flow cytometry data. This is accomplished through a
data model very close to that adopted by other Bioconductor packages
used for gene expression analysis and thus familiar to most
Bioconductor users. Similar to any other flow tool, the basic unit of
organization in flowCore is a collection of events. We call the
structure that hold this data a flowFrame. All relevant information
for the collection are stored along with the matrix of raw data values
and can be accessed programmatically. Most prominently, these are
descriptors of the stains used in the experiment and the respective
measurement channels, information about compensation performed at the
instrument side and any additional keywords the user deems to be
important to annotate the data. A number of quality checks are
performed during the creation of a flowFrame to ensure data
integrity. There is an extended interface that allows for the
standardized interaction with the flow data, thus enabling the
development of more specialized tools for specific tasks of the data
analysis process.

\subsection{flowSet}
In high-throughput flow cytometry, many of the analysis tasks need to
be performed consistently across multiple flow measurements, hence we
introduce the concept of a collection of flowFrames called a
flowSet. flowSets store the relevant information associated with each
individual frame such as descriptions of the samples or the treatment
to which a sample was subjected, and they manage the consistent
application of operations on the individual flowFrames. The burden of
keeping score of the annotation is shifted from the user to the
infrastructure, thus reducing the risk of mixups and errors that are
all too common when dealing with complex data in the ususal way of
more or less ad hoc spreadsheets that lack a controlled structure. In
general, there is no native ordering to a flowSet, which in a way
reflects the still predominant application of tube-based experiments
rather than experiments following a plate-design. However, the flowSet
structure can be readily extended to also accommodate such spacial
information \textit{(reference)}

@@

\section{Standard flow operations}
The basic operations in flow cytometry analysis are almost always the
same: the data need to be compensated (if not already done on the
instrument side), often transformed and sup-population of interest
need to be selected based on a set of (predominantly sequential)
gates. All software solutions for flow data analysis offer more or
less elaborate support for these kinds of operation, most of them in
an interactive, GUI-driven spirit. The approach we have taken in
flowCore is to abstractly describe these operations and build a set of
tools to perform them on both flowFrames as well as on entire
flowSets. While transformation and to a certain extent also
compensation are fairly routine operations with only limited potential
for improvement, being able to implement new methodology for gating of
flow cytometry data and extending the capabilities of flowCore through
the usual ways of object oriented programming is a feature, that
clearly sets our framework apart from all the other flow toolsets
around. By factoring out as much of the bookkeeping as possible,
authors can focus on the actual operations rather then having to deal
with the tedious details of data integration and access. Third party
methods implemented by others can act on their own as first-class
citizens in the analysis, without breaking the workflow or the basic
infrastructure. This design allows for the straight forward extension
of flowCore's capabilities and already fostered the development of a
number of valuable improvements. \textit{(cite stuff here)}

@@

\subsection{gating}
In flowCore, gates are represented as individual objects that can be
extended in an object oriented manner. All basic gate types such as
rectangular gates, ellipses and polygons are implemented as part of
the base framework. In addition, we introduce the notion of
data-driven gates, for which the necessary parameters are computed
based on the properties of the underlying data, for instance
distribution or density estimates. The ability to programmatically
access gates is a prerequisite for automated or semi-automated
gating. By utilizing a unified interface for all different types of
gates, the user is able to subset data sets as well as create summary
statistics, for instance of proportions of events falling in single
gates or combinations of multiple gates. The definition of gates in
flowCore follows the Gating-ML CR, thus any flowCore
gating strategy can be reproduced by any other software that also
adheres to the standard.

\section{Related flow packages}
In addition to the flowCore package that offers basic infrastructure
we have implemented a range of additional Bioconductor packages that
are dedicated to more specific tasks of flow cytometry data
analysis. The flowViz package provides sophisticated data
visualization tools that make use of multivariate trellis
plotting. These functions can be used to quickly generate customized
plots for extended cytometry data sets for both direct data inspection
and quality control. The flowQ package offers more advanced quality
assurance (QA) methodology and a framework to create interactive
reports of QA results. Most of the flowUtils content deals with data
import and output from and to flowCore and the flow-cytometry specific
standard markup languages. Finally, flowStats provides statistical
methods that are relevant in the context of flow cytometry data
analysis.
